---
title: "RNA-seq analysis in R: Mouse mammary RNA-Seq data"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<<<<<<< HEAD
# Mouse Mammary Gland Dataset
The data for this R markdown file comes from a Nature Cell Biology paper, EGF-mediated induction of Mcl-1 at the switch to lactation is essential for alveolar cell survival (Fu et al. 2015). Both the raw data (sequence reads) and processed data (counts) can be downloaded from Gene Expression Omnibus database (GEO) under accession number GSE60450.

This study examines the expression profiles of basal stem-cell enriched cells (B) and committed luminal cells (L) in the mammary gland of virgin, pregnant and lactating mice. Six groups are present, with one for each combination of cell type and mouse status. Each group contains two biological replicates. We will first use the counts file as a starting point for our analysis. This data has already been aligned to the mouse genome. The command line tool featureCounts (Liao, Smyth, and Shi 2014) was used to count reads mapped to mouse genes from Refseq annotation (see the paper for details).

=======
>>>>>>> 5902aaf6cfd38206a674810f6c76bce380952fc5
# Installing Packages and Loading Libraries
```{r}
if (!requireNamespace("BiocManager"))
    install.packages("BiocManager")
BiocManager::install(c("limma", "edgeR", "Glimma", "org.Mm.eg.db", "gplots", "RColorBrewer", "NMF", "BiasedUrn"))
```

```{r}
library(edgeR)
library(limma)
library(Glimma)
library(org.Mm.eg.db)
library(gplots)
library(RColorBrewer)
library(NMF)
```

<<<<<<< HEAD
# Data Import
```{r}
# Read the data into R
seqdata <- read.delim("data/GSE60450_LactationGenewiseCounts.txt", stringsAsFactors = FALSE)
# Read the sample information into R
sampleinfo <- read.delim("data/SampleInfo.txt", stringsAsFactors = TRUE)
```

=======
>>>>>>> 5902aaf6cfd38206a674810f6c76bce380952fc5

#Format the data
```{r}

# Remove first two columns from seqdata
countdata <- seqdata[,-(1:2)]
seqdata
# using substr, characters starting at position 1 and stopping at position 7 of the colnames are extracted
colnames(countdata) <- substr(colnames(countdata),start=1,stop=7)

# Store EntrezGeneID as rownames
rownames(countdata) <- seqdata[,1]

# Look at the output
head(countdata)

# the column names are now the same as SampleName in the sampleinfo file.  it means our sample information in sampleinfo is in the same order as the columns in countdata.
table(colnames(countdata)==sampleinfo$SampleName)

```

# Convert counts to DGEList Object
DGEList object is used by edgeR to store count data and has a number of slots for storing various parameters about the data.
```{r}
y <- DGEList(countdata)
y

# See what slots are stored in y
names(y)

# Library size information is stored in the samples slot
y$samples


# store the groups for the samples in the DGEList object.
group <- paste(sampleinfo$CellType,sampleinfo$Status,sep=".")
# Take a look
group

# Convert to factor
group <- factor(group)
# Take another look.
group

# Add the group information into the DGEList
y$samples$group <- group
y$samples
```
# Adding Annotations
The Entrez gene id is the only annotation we can see which is not bery helpful so we add some annotation info using org.Mm.eg.db package.
```{r}
#columns function shows what info is available in the database
columns(org.Mm.eg.db)

#a separate dataframe containing gene symbols and gene name along with entrezid is formed
ann <- select(org.Mm.eg.db,keys=rownames(y$counts),columns=c("ENTREZID","SYMBOL","GENENAME"))
head(ann)

#double check whether the ENTRZID column matches exactly to our y$counts row 
table(ann$ENTREZID==rownames(y$counts))

#the annotation information is then slotted into genes slot of y
y$genes <- ann
```

# Filtering Lowly Expressed Genes
Genes with low counts provide little evidence for differential expression and interfere with the statistical approximations used in the pipeline. To filter out lowly expressed genes, we choose to retain genes if they are expressed at a counts-per-million above 0.5 in at least two samples.
```{r}
# obtaining CPMs 
myCPM <- cpm(countdata)
# output
head(myCPM)
```

```{r}
# checking values in myCPM greater than 0.5, by producing a logical matrix of TRUEs and FALSEs
thresh <- myCPM > 0.5
head(thresh)
```

```{r}
# summary of how many TRUEs are in each row
table(rowSums((thresh)))
```

```{r}
#keeping genes that have at least 2 TRUEs in each row of thresh
keep <- rowSums(thresh) >= 2
summary(keep)
```
```{r}
# checking whether the threshold of 0.5 correspond to a count of 10-15
plot(myCPM[,1], countdata[,1], ylim =c(0,50), xlim = c(0,3))
# adding vertical line at 0.5 CPM and horizontal lune at 10
abline(v=0.5, h = 10, col = "blue")
```
```{r}
# after we have checked the filtering method, we will filter the DGEList object
y <- y[keep, keep.lib.sizes = FALSE]
```

#Quality Control
After we have removed lowly expressed genes an have the counts stores in a DGEList object, we can draw some plots to check the quality of the data

## Library Size and Distribution Plots
```{r}
# checking how many reads we have for each sample
y$samples$lib.size
```
We can plot the library sizes as a barplot to see whether there are any major discrepancies between the samples more easily
### Barplot
```{r}
# names argument tels the barplot to use the sample names on the x-axis, las argument rotates the axis names
barplot(y$samples$lib.size/1e06, names = colnames(y), las = 2, cex.name = 0.75, ann = FALSE)
mtext(side = 1, text = "Samples", line = 4)
mtext(side = 2, text = "Library Size (in millions)", line = 3)
title("Barplot of Library Sizes")
```
### Boxplot
Since count data is not normally distributed, we need to log the raw counts and use box plots to check their distribution
```{r}
# get log2 counts per million
logcounts <- cpm(y, log = TRUE)
# boxplot
boxplot(logcounts, xlab = "", ylab = "log2 counts per million", las = 2)
# adding a blue horizontal line that corresponds to the median logcpm
abline(h=median(logcounts), col = "blue")
# title
title("Boxplots of logCPMs C(unnormalised)")
```
## Multidimensional Scaling Plots
MDSplots are one of the most important plots to anlayse RNA-Seq data. MDSplot is a visualization of PCA, which determines the greatest sources of variation in the data.
```{r}
plotMDS(y)
```
```{r}
# how many cell types and in what order are they stored
levels(sampleinfo$CellType)
```

```{r}
#purple for basal and orange for luminal
col.cell <- c("purple", "orange")[sampleinfo$CellType]
data.frame(sampleinfo$CellType,col.cell)
```
```{r}
# Similarly for status
levels(sampleinfo$Status)

# blue red and black for lactate, pregnant and virgin
col.status <- c("blue","red","black")[sampleinfo$Status]
col.status
```

```{r}
# specifying option to let us plot two plots side-by-side
par(mfrow= c(1,2))

# redoing the MDS with cell type colouring
plotMDS(y, col = col.cell)
legend("topleft", fill = c("purple", "orange"),legend = levels(sampleinfo$CellType))
# Adding a title
title("Cell type")

# redoing the MDS with status colouring
plotMDS(y,col=col.status)
legend("topleft",fill=c("blue","red","black"),legend=levels(sampleinfo$Status),cex=0.8)
title("Status")


```
```{r}
# Sampleinfo Correction
sampleinfo <- read.delim("data/SampleInfo_Corrected.txt", stringsAsFactors = TRUE)
sampleinfo
group <- factor(paste(sampleinfo$CellType,sampleinfo$Status,sep="."))
y$samples$group <- group
```
```{r}
# Redoing the MDSplot with corrected information
par(mfrow=c(1,2))
col.cell <- c("purple","orange")[sampleinfo$CellType]
col.status <- c("blue","red","black")[sampleinfo$Status]
plotMDS(y,col=col.cell)
legend("topleft",fill=c("purple","orange"),legend=levels(sampleinfo$CellType))
title("Cell type")
plotMDS(y,col=col.status)
legend("topleft",fill=c("blue","red","black"),legend=levels(sampleinfo$Status),cex=0.8)
title("Status")
```
## Hierarchical Clustering with Heatmaps
Heatmaps help in the visualization of hierarchical clustering of the samples. Here, heatmap.2 function from the gplots package is used which calculates a matrix of euclidean distances from the logCPM (logcounts object) for the 500 most variable genes.
Now, selecting data for the 500 most variable genes and plotting the heatmap
```{r}
# estimating the variance for each row in the logcounts matrix
var_genes <- apply(logcounts, 1, var)
head(var_genes)
```
```{r}
# getting gene names for the top 500 most variable genes
select_var <- names(sort(var_genes, decreasing = TRUE))[1:500]
head(select_var)
```
```{r}
# subsetting logcounts matrix
highly_variable_lcpm <- logcounts[select_var,]
dim(highly_variable_lcpm)
head(highly_variable_lcpm)
```
```{r}
# getting colors
mypalette <- brewer.pal(11,"RdYlBu")
morecols <- colorRampPalette(mypalette)
# setting up color vector for celltype variable
col.cell <- c("purple", "orange")[sampleinfo$CellType]
# plot the heatmap
heatmap.2(highly_variable_lcpm,col=rev(morecols(50)),trace="none", main="Top 500 most variable genes across samples",ColSideColors=col.cell,scale="row")
```
```{r}
# more annotations, like colors for cell type and status, the aheatmap function can be used. which columns to use can be specified using sampleinfo in annCol= which adds colored bars.
mypalette <- brewer.pal(11,"RdYlBu")
morecols <- colorRampPalette(mypalette)
aheatmap(highly_variable_lcpm,col=rev(morecols(50)),main="Top 500 most variable genes across samples",annCol=sampleinfo[, 3:4],labCol=group, scale="row")
```
# Normalization for Composition Bias
Normalization needs to be done to eliminate composition biases between libraries
```{r}
# Applying normalisation to DGEList object
y <- calcNormFactors(y)
```
The calcNormFactors function updates the normalisation factors in the DGEList. A value below one one indicates that the library size will be scaled down, as there is more suppression (i.e., composition bias) in that library relative to the other libraries. This is also equivalent to scaling the counts upwards in that sample. Conversely, a factor above one scales up the library size and is equivalent to downscaling the counts. *Add MD plots later
```{r}
y$samples
```
# Differential Expression with limma-voom
Since we have normalised the data and the quality looks good, we can start testing for differentially expressed genes. For this, we used the limma package which offers the voom function , which transforms the read counts into logCPMs while taking into account the mean-variance relationship in the data. After vooming, we can apply a linear model to the voom transformed data to test for differentially expressed genes, using standard limma commands.

## Creating the design matrix








